
---
title: "Lab03: Data Cleaning with R"
author: "TECK Squad"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview & Goals

In this lab, you will **continue working with the web advertising bids data** to:

- Systematically **detect** and **repair** some other data quality problems using those ideas.
- Create a **cleaned version** of the dataset (e.g., `bids_clean`).
- Document your work in a **data cleaning log**.

By the end, you should have:

1. A data set with the majority of the insidious issues corrected.
2. A short narrative (or table) explaining **what you fixed** and **how**.

> *Work in the **same repository** as previous labs. Update your Jira Kanban board with new cards for this lab and keep committing your progress with clear messages.*

# Setup

## Packages

Minimally, you will need these packages:

```{r loadpackages}
library(tidyverse)
library(arrow)
library(dplyr)
library(stringr)
```

- `tidyverse` (especially `dplyr`, `ggplot2`, `readr`, `stringr`)
- `stringr` to modify character strings
- `lubridate` for date–time parsing
- `arrow` if you are reading/writing Parquet files

> **Task:** In a code chunk, load the packages you need.

## Load the messy bids data

Use the **same dataset** as in the earlier lab (e.g., a Parquet file in your `data/` folder).

> **Tasks:**
>
> 1. Read the messy data into R (e.g., using `arrow::read_parquet()` or `readr` functions, depending on your file).
> 2. Use `dplyr::glimpse()` and/or `summary()` to remind yourselves what variables are present and roughly what they look like.
> 3. Create a **working copy** of the data so that you can always go back to the original if needed.

You may want to keep the original as `bids_raw` and your working copy as `bids`.

```{r Read Data}
bids <- read_parquet("../data/bids_data_vDTR.parquet")

```


# Quick Orientation Check (10–15 minutes)

Using your prior lab work and Data Cleaning lectures:

1. In **2–4 sentences**, describe **what each row** in this dataset represents.
2. Identify **one or two variables** that you think should be:
   - **Mandatory** (must not be missing), and
   - Possibly a **unique key** or part of a composite key.
3. Pick **two variables** and, for each, assign at least one **data quality dimension** (e.g., completeness, validity, consistency, uniqueness, cross-field consistency).

> **Write your answers directly in this Rmd in a text section (not code).**

# Fixing Numeric & Price Issues (30–40 minutes)

Focus on the **PRICE** variable (and other numeric variables if needed).

You should investigate:

- **Type issues** (e.g., stored as character instead of numeric),
- **Impossible values** (negative values, weird “huge” values),
- **Sentinel codes** (e.g., `-999`),
- Extra non-numeric characters.

## Suggested tools

- `class()`, `unique()`, `dplyr::count()`, `summary()`
- `readr::parse_number()`
- `as.numeric()`
- `dplyr::mutate()`, `dplyr::case_when()`
- `is.na()`, `sum(is.na(...))`
- `quantile()` to inspect distributions

## Inspect PRICE

> **Tasks:**
>
> 1. Check the **type** of `PRICE` and inspect a sample of values (e.g., using `sample()` or `head()`).

```{r}
# Your code here

```

> 2. In short sentences, describe what you see:
>    - Is `PRICE` numeric or character?
>    - Do you see obvious anomalies or strange values?


## Create a clean numeric version of PRICE

Create a **cleaned numeric version** (e.g., `PRICE_clean`) that:

- Extracts numeric content from strings (if necessary),
- Converts to numeric,
- Keeps track of values that cannot be converted (they will become `NA`).

> **Tasks:**
>
> 1. Use one or more of these:
>    - `readr::parse_number()`
>    - `stringr::str_remove_all()`
>    - `as.numeric()`
>
>    To learn about their use, type `?` before either of them (e.g., `?readr::parse_number()`)

```{r}
# Your code here

```

> 2. Count how many `NA`s you get in `PRICE_clean`. Explain why some values might be `NA` after parsing.


```{r}
# Your code here

```


## Define and apply rules for impossible values

Now decide how to handle:

- Negative prices,
- Very large prices that are implausible,
- Sentinel codes (like `-999`).

> **Tasks:**
>
> 1. Use `dplyr::filter()` and summaries (`summary()`, `quantile()`) to identify at least **two kinds** of “bad” price values.


```{r}
# Your code here

```

> 2. Write down your **rules** for how to handle each type (e.g., set to `NA`, drop rows, cap values).
> 3. Implement those rules in a new variable (e.g., `PRICE_final`) using `dplyr::mutate()` and `dplyr::case_when()`.

```{r}
# Your code here

```

> 4. Briefly describe your rules in text and justify them using the **validity** and **completeness** dimensions of data quality.


# Cleaning Categorical & String Fields (30–40 minutes)

Now work on **categorical** and **string** variables, such as:

- `DEVICE_GEO_REGION`
- `DEVICE_GEO_ZIP`
- `RESPONSE_TIME`

Remember: **this dataset only includes bids from Oregon**, so a correct region should be coded as `"OR"`. Other values in `DEVICE_GEO_REGION` represent **modified/mangled versions** of `"OR"`.

Helpful string tools:

- `stringr::str_trim()`, `stringr::str_to_upper()`, `stringr::str_replace()`
- `stringr::str_detect()`, `stringr::str_pad()`, `stringr::str_length()`
- `dplyr::count()` to see category frequencies

## DEVICE_GEO_REGION: standardizing Oregon codes

> **Tasks:**
>
> 1. Use `dplyr::count(DEVICE_GEO_REGION, sort = TRUE)` to list distinct region values and their frequencies.

```{r}
# Count all unique region values and sort them by frequency
bids %>%
  dplyr::count(DEVICE_GEO_REGION, sort = TRUE)


```

> 2. Identify **all values** that clearly correspond to **Oregon**, just coded inconsistently (e.g., extra punctuation, different case, misspellings).


```{r}
# Inspect simplified versions of the region strings to identify Oregon variants
bids <- bids %>%
  dplyr::mutate(
    region_trim  = stringr::str_trim(DEVICE_GEO_REGION),          # remove extra spaces
    region_upper = stringr::str_to_upper(region_trim),            # convert to uppercase
    region_clean = stringr::str_replace_all(region_upper, "[^A-Z]", "")  # keep letters only
  )

bids %>%
  dplyr::count(region_clean, sort = TRUE)

```

> 3. Define a **standard representation** (`"OR"`) and create a cleaned version (e.g., `DEVICE_GEO_REGION_clean`) where:
>    - All Oregon variants are standardized to `"OR"`.
>    - Other values (if any) are treated according to a rule you decide (e.g., set to `NA` if they cannot be interpreted as Oregon).


```{r}
bids <- bids %>%
  mutate(
  # Standardize all Oregon variants to "OR"
    DEVICE_GEO_REGION_clean = dplyr::case_when(
      region_clean %in% c("OR", "ORE", "OREG", "OREGON", "XOR") ~ "OR",
      TRUE ~ NA_character_
    )
  )

# Check the cleaned region values and their frequencies
bids %>%
  dplyr::count(DEVICE_GEO_REGION_clean, sort = TRUE)

```

> 4. In **2–3 sentences**, explain:
>    - How you detected inconsistent codes,
>    - The **set-membership rule** you are enforcing (valid region codes for this dataset).

We used dplyr count() to inspect all distinct DEVICE_GEO_REGION values and identify inconsistent Oregon codes. After trimming whitespace and converting all values to uppercase, we standardized all Oregon variants (“OR”, “ORE”, “OREG”, “OREGON”, and “XOR”) to the clean code “OR”, while treating any other unexpected patterns as invalid and setting them to NA.

## DEVICE_GEO_ZIP: formats and sentinels

Treat ZIP codes as **strings**, not numbers.

> **Tasks:**
>
> 1. Convert ZIP codes to character and use `dplyr::count()` to inspect common values.


```{r}
# List the most common ZIP values (already stored as character)
bids %>%
  count(DEVICE_GEO_ZIP, sort = TRUE)

```

> 2. Identify suspicious ZIP values:
>    - Too short / too long,
>    - Obvious sentinel codes (e.g., `-999`, `9999`),
>    - Non-digit characters.

```{r}
# Helper table for ZIP checks (used in Tasks 2–4)
bids_zip_check <- bids %>%
  mutate(
    # 1) Normalize ZIP: trim spaces
    zip_trim  = stringr::str_trim(zip_raw),

    # 2) Length of the full ZIP string
    zip_length = stringr::str_length(zip_trim),

    # 3) Sentinel codes we want to treat as invalid
    is_sentinel = zip_trim %in% c("-999"),

    # 4) Does the ZIP start with a negative sign?
    starts_negative = stringr::str_starts(zip_trim, "-"),

    # 5) First 5 characters (candidate 5-digit ZIP)
    zip5 = stringr::str_sub(zip_trim, 1, 5),

    # 6) Check that the first 5 chars are exactly 5 digits
    zip5_is_digits = stringr::str_detect(zip5, "^[0-9]{5}$"),

    # 7) Convert those 5 digits to integer (for range checks)
    zip5_int = dplyr::if_else(zip5_is_digits, as.integer(zip5), NA_integer_),

    # 8) Oregon range check: 97001–97920
    out_of_or_range = dplyr::if_else(
      !is.na(zip5_int) & (zip5_int < 97001L | zip5_int > 97920L),
      TRUE, FALSE
    ),

    # 9) Any non-digit characters anywhere in the trimmed ZIP?
    has_nondigit = !stringr::str_detect(zip_trim, "^[0-9-]+$")
  )

# Task 2: show only *suspicious* ZIP patterns
suspicious_zips <- bids_zip_check %>%
  filter(
    zip_length < 5 | zip_length > 10 |   # too short / too long
      is_sentinel |                      # sentinel values
      starts_negative |                  # negative ZIPs
      has_nondigit |                     # non-digit characters
      out_of_or_range                    # outside Oregon range
  ) %>%
  count(zip_trim, sort = TRUE)

suspicious_zips

```

> 3. Decide on a **pattern** for valid ZIP codes (e.g., Oregon 5-digit ZIPs).
>    - Consider using `stringr::str_length()`, `stringr::str_detect()` with a regular expression pattern, and `stringr::str_pad()` if necessary.

```{r}

# Use the helper table from Task 2
bids_zip_check <- bids_zip_check %>%
  mutate(
    is_valid_oregon_zip =
      !is.na(zip5_int) &       # valid 5-digit ZIP
      !is_sentinel &           # not -999
      !starts_negative &       # does not start with "-"
      !has_nondigit &          # only digits
      !out_of_or_range         # inside Oregon range
  )

bids_zip_check %>% count(is_valid_oregon_zip)


```

> 4. Create a cleaned version (e.g., `DEVICE_GEO_ZIP_clean`) that:
>    - Trims whitespace,
>    - Converts sentinel codes to `NA`,
>    - Enforces your chosen ZIP pattern.

```{r}
# Final cleaned ZIP using the helper table (no repeated cleaning)
bids <- bids_zip_check %>%
  mutate(
    DEVICE_GEO_ZIP_clean = if_else(
      is_valid_oregon_zip,
      zip5,           # keep the first 5 digits
      NA_character_   # otherwise NA
    )
  )

# Look at the cleaned ZIP distribution to confirm the result
bids %>%
  dplyr::count(DEVICE_GEO_ZIP_clean, sort = TRUE)

```

> 5. In a short paragraph, describe your rule and why it is reasonable given the context (Oregon-only data).

Our rule defines a valid ZIP code as a clean 5-digit Oregon ZIP between 97001 and 97920. Because the data only contains Oregon bids, any value outside this range cannot be real. After reviewing the raw ZIPs, we flagged issues such as sentinel values (-999), negative signs, non-digit characters, and incorrect lengths. We treated “-999” as NA because it is a negative, impossible ZIP value that falls outside Oregon’s valid bounds and clearly acts as a sentinel placeholder. After extracting the first five digits, we kept only values that passed all checks and set the rest to NA.

## RESPONSE_TIME: prefixes and extra characters

The `RESPONSE_TIME` variable may include:

- A **text prefix** (e.g., “RESPONSE_TIME:” or typo’d variants),
- Extra characters at the end (e.g., punctuation),
- Numeric content representing a duration.

> **Tasks:**
>
> 1. Use a small `sample()` of `RESPONSE_TIME` to inspect the raw patterns.

```{r}
# Take a small random sample of the raw RESPONSE_TIME values to inspect the patterns
sample(bids$RESPONSE_TIME, size = 20)


```

> 2. Write a rule for how to:
>    - Remove any “RESPONSE_TIME:” prefixes (including likely typos),
>    - Strip extra non-numeric characters from the end,
>    - Convert the remaining content to a numeric type.

To clean the RESPONSE_TIME field, my rule is to first remove any text prefix at the beginning of the value, including typos or variations of “RESPONSE_TIME”.  I then remove any additional non-numeric characters that might be present at the end of the string.I transform the cleaned value into a numeric type so that it can be used for analysis once only the digits are left.

> 3. Implement your rule in a cleaned variable (e.g., `RESPONSE_TIME_clean`) using `stringr::str_replace()` or related functions and `as.numeric()`.

```{r}
# Clean RESPONSE_TIME
bids <- bids %>%
  mutate(
    # 1) Remove ANY prefix (correct or typo) = remove all characters until first digit
    rt_no_prefix = stringr::str_replace(RESPONSE_TIME, "^[^0-9]+", ""),

    # 2) Remove any trailing non-numeric characters
    rt_digits_only = stringr::str_replace(rt_no_prefix, "[^0-9]+$", ""),

    # 3) Convert to numeric
    RESPONSE_TIME_clean = as.numeric(rt_digits_only)
  )

# Check result
bids %>% count(RESPONSE_TIME_clean, sort = TRUE)
summary(bids$RESPONSE_TIME_clean)


```

> 4. Count how many `NA`s you get in `RESPONSE_TIME_clean` and briefly interpret what those `NA`s mean.

```{r}
# Count missing values in RESPONSE_TIME_clean
bids %>%
  dplyr::summarise(
    num_NA = sum(is.na(RESPONSE_TIME_clean)),
    total  = dplyr::n()
  )


```

> 5. Connect this to the idea of **regular-expression patterns** as a tool for enforcing validity.

By outlining precisely what a "correct" value should be, regular-expression patterns aid in enforcing validity.In this task, we eliminated invalid prefixes and trailing characters by using regex rules to extract only the numeric portion of each RESPONSE_TIME entry.By enforcing this pattern, we improve data quality and consistency by ensuring that only correctly formatted response times are retained and that any values that do not match the expected numeric structure become NA.

# Date–Time Cleaning: TIMESTAMP (20–30 minutes)

The `TIMESTAMP` field records when a bid occurred, but the **format is inconsistent** (for example, some dates may use `-` and others `/`).

Useful tools:

- `as.character()`
- `stringr::str_detect()`
- `lubridate::parse_date_time()`, `lubridate::ymd_hms()`, `lubridate::mdy_hms()`

## Inspect TIMESTAMP patterns

> **Tasks:**
>
> 1. Look at a small random sample of `TIMESTAMP` values.

```{r}
sample(bids$TIMESTAMP)
```

> 2. Use `stringr::str_detect()` to count how many timestamps use `"-"` versus `"/"` in the date.

```{r}
stringr::str_detect(bids$TIMESTAMP, "-") %>%
  sum()
stringr::str_detect(bids$TIMESTAMP, "/") %>%
  sum()
```

> 3. Summarize in 1–2 sentences what you observe.

There are two obvious issues so far: NA values in the date portion of the variable, and some entries are listed with "/" rather than "-". There are 1157 entries where "/" is used.

## Parse TIMESTAMP into a POSIXct variable

> **Tasks:**
>
> 1. Use `lubridate::parse_date_time()` (or similar) to create `TIMESTAMP_clean`:
>    - Allow for multiple possible formats (e.g., one order for `ymd HMS` and one for `mdy HMS` to see examples scroll paste `?lubridate::parse_date_time()` in your console and scroll all the way to the bottom of the function's help tab),
>    - Specify a time zone (e.g., `"UTC"`).

```{r TIMESTAMP Parse}
TIMESTAMP_clean <- lubridate::parse_date_time(bids$TIMESTAMP, c("ymd HMS", "mdy HMS"), tz = "UTC")

```

> 2. Count how many `NA`s result from parsing.

```{r TIMESTAMP NAs}
sum(is.na(TIMESTAMP_clean))
```

> 3. Use `summary()` or `range()` to examine the minimum and maximum of `TIMESTAMP_clean`.

```{r TIMESTAMP summary}
summary(TIMESTAMP_clean)
```

> 4. Briefly comment on:
>    - Whether the time range seems reasonable,
>    - What kinds of entries failed to parse (if any).

The shown range for the TIMESTAMP variable seems reasonable, given that we only have about a day's worth of data. There were entries that had a given time, but the date was labeled as NA. These seem to be the entries that did not parse into the POSIXct variable. Though the date is NA, we know the range is within a day. There may be possibility to recover those date values by referencing the given time and assigning the date associated with it.

# Keys, Duplicates, and Out-of-Range Values (20–30 minutes)

## Candidate keys and duplicates

Using your earlier reasoning:

> **Tasks:**
>
> 1. Choose one variable or a combination of variables that you **believe should uniquely identify** a row (a key).
> 2. Use `dplyr::count()` to check whether that key is truly unique or if there are duplicates.

```{r}
# Your code here

```

> 3. Check for **full-row duplicates** using `dplyr::distinct()` or by counting occurrences across all columns.

```{r}
# Your code here

```

> 4. Decide on a rule for handling duplicates:
>    - Drop exact duplicates?
>    - Keep first occurrence and drop the rest?
>    - Something else?
> 5. Implement your rule and create a de-duplicated dataset (e.g., `bids_nodup`).

```{r}
# Your code here

```

> 6. In 2–3 sentences, explain your choice and connect it to the **uniqueness** dimension of data quality.

## Out-of-range latitude/longitude (optional but recommended)

Look at `DEVICE_GEO_LAT` and `DEVICE_GEO_LONG`.

> **Tasks:**
>
> 1. Use `summary()` or `dplyr::summarise()` to get the minimum and maximum latitude and longitude.

```{r}
# Your code here

```

> 2. Compare the ranges to what you know about **Oregon’s** geographic coordinates (you can approximate based on general US geography; no need for exact boundaries).

```{r}
# Your code here

```

> 3. Decide which coordinates are clearly **implausible**.

```{r}
# Your code here

```

> 4. Define rules to create cleaned versions (e.g., `DEVICE_GEO_LAT_clean`, `DEVICE_GEO_LONG_clean`) where:
>    - Plausible values are kept,
>    - Implausible values are set to `NA` or corresponding rows are removed.

```{r}
# Your code here

```

> 5. Document the rules you used and why.

# Creating a Final Cleaned Dataset (10–15 minutes)

Now pull everything together.

> **Tasks:**
>
> 1. Create a final cleaned dataset (e.g., `bids_clean`) that:
>    - Uses your cleaned variables (`PRICE_final`, `DEVICE_GEO_REGION_clean`, `DEVICE_GEO_ZIP_clean`, `RESPONSE_TIME_clean`, `TIMESTAMP_clean`, cleaned lat/long, etc.),
>    - Drops unnecessary intermediate columns (e.g., `_raw`, helper strings) that you don’t need for analysis.

```{r}
# Your code here

```

> 2. Check the structure of `bids_clean` with `dplyr::glimpse()`.

```{r}
# Your code here

```

> 3. Optionally, write the cleaned dataset to disk (for your own future use) using something like `save()` (see `?save` for its use and an example) to save in `.RData` format **but do not commit the cleaned data file** to the remote repository.

```{r}
# Your code here (optional)

```

> 4. In a short paragraph, describe what your final cleaned dataset now represents and what major issues have been addressed.

# Data Cleaning Log & Reflection (10–15 minutes)

## Data cleaning log (table)

Create a succinct table summarizing the main problems and fixes.

Use the template below and fill it in with your team’s work. **One row (PRICE) is completed as an example; you should complete the remaining rows yourselves.**

| Variable | Problem Detected (brief) | Data Quality Dimension(s) | Cleaning Action (brief) |
|----------|-------------------------|---------------------------|--------------------------|
| Variable           | Problem Detected (brief)                                                | Data Quality Dimension(s) | Cleaning Action (brief)                                                         |
|-------------------|--------------------------------------------------------------------------|----------------------------|--------------------------------------------------------------------------------|
| PRICE             | Includes non-numeric characters                                          | Validity                   | Extract numeric portion and convert to numeric                                  |
| DEVICE_GEO_REGION | Inconsistent forms ("OREGON", "XOR", lowercase, extra spaces/punctuation) | Consistency / Validity     | Trim, uppercase, remove punctuation; map "OR", "OREGON", "XOR" → "OR"; others → NA |
| DEVICE_GEO_ZIP    | Sentinel codes (-999, 9999, 00000), non-digit chars, wrong length       | Validity / Accuracy        | Remove sentinels; trim; convert to character; pad to 5 digits; keep valid Oregon ZIPs (97xxx) |
| RESPONSE_TIME     | Prefix "RESPONSE_TIME:" and trailing non-digit characters                | Validity                   | Remove prefix; strip non-digit endings; convert to numeric; invalid → NA        |
| TIMESTAMP         | *(team describes)*                                                       | *(team selects)*            | *(team describes)*                                                              |
| Other variable(s) | *(optional)*                                                             | *(optional)*                | *(optional)*                                                                    |

## Reflection (5–8 sentences)

Write a short reflection addressing:

- Which data cleaning tasks were the most **time-consuming** or **surprising**?
- Which functions or workflows (e.g., `mutate()`, `case_when()`, `str_*()`, `parse_date_time()`, `distinct()`) were most valuable?
- One thing you will do **differently** the next time you receive a new messy dataset.
- Whether/how you used generative AI for this lab and how it affected your understanding.

**Khloud**
Cleaning this dataset was more time-consuming than I expected, especially the string-based tasks such as standardizing region codes and validating ZIP formats. Many values were inconsistent, which made it surprising how easily fields can become messy in real-world data. Functions like mutate(), case_when(), str_trim(), str_detect(), and distinct() were especially valuable for organizing and transforming variables step by step. If I receive a similar dataset in the future, I would first create a clear plan for handling string patterns and sentinel values before writing any code. I also used generative AI to help me understand errors, write some of the cleaning code, and correct mistakes. This support made the workflow more efficient and improved my understanding of data-cleaning techniques.
